---
title: "R Notebook"
output: html_notebook
---

```{r, warning=F, error=F, message=F}

# Packages

rm(list=ls())

require(tidyverse)
require(pgirmess)
require(ncf)
require(geoR)
require(ggplot2)
require(terra)

```


# Background

Researchers set up a square plot with a dimension of 130 m on each side. Within the plot they established blue vane traps in a 5 m grid pattern to sample bee abundance. Every day for 10 days the researchers emptied the blue vane traps and recorded the number of bees collected. After the 10-day sampling period, they calculated the mean daily number of bees captured in the trap each day.

# Challenge 1 (2 points)

Import the blueVanes.csv file from GitHub. Plot the mean bee abundance as a function of the x and y coordinates. Upon initial inspection, does there appear to be spatial dependence in the bee response data?

```{r}
# blueVanes

blueVanes = read.csv("https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week6/blueVanes.csv")
head(blueVanes)

ggplot(blueVanes, aes(x=x, y=y, color=meanBee))+
  geom_point(size = 5)+
  theme_bw()+
  theme(panel.grid=element_blank())+
  scale_color_gradient(low='blue', high='red')
```

Answer the question here.

I think that there does appear to be some weak spatial dependence in the abundance data (at least, it is weak compared to the example in the lab). I think that there is spatial dependence in the abundance data because there appears to be more mid-abundance values by each other than there would be in a truly random distribution. It appears to be weak, however, because the extreme values are isolated and the values are not grouped as obviously as the example in lab.

# Challenge 2 (3 points)

Empirically test for evidence of spatial dependence in bee abundance by first building a corelogram, then building a semi-variogram. Be sure to provide an envelope for assessing whether the values in each figure are statistically different from what would be expected if the bee values were randomly distributed. Describe the patterns you see, and your interpretation with regards to the spatial dependence in mean bee abundance. Be sure to identify the maximum distance at which you should interpret your figures (2/3 of the maximum distance between points).

```{r}
# Distances

coords = cbind(blueVanes$x, blueVanes$y)
colnames(coords) = c('x', 'y')
distMat = as.matrix(dist(coords))

maxDist = (2/3) * max(distMat)

# Correlogram - MC permutation (if our values were randomly distributed)

correlogNcf = ncf::correlog(x = blueVanes$x, y=blueVanes$y, z=blueVanes$meanBee, increment=10, resamp=100)

tmp = cbind(correlogNcf$n,
            correlogNcf$mean.of.class,
            correlogNcf$correlation,
            correlogNcf$p) %>% 
  data.frame() %>% 
  set_names(c('n', 'dist', 'corr', 'p')) %>% 
  mutate(significant = ifelse(p < 0.05, 'yes', 'no'))

ggplot(tmp, aes(x=dist, y=corr))+
  geom_line()+
  geom_point(aes(color=significant, size=n))+
  scale_color_manual(values=c('black', 'red'))+
  theme_bw()+
  theme(panel.grid=element_blank())+
  geom_hline(yintercept=0, linetype='dashed')+
  geom_vline(xintercept=maxDist, color='red')+
  scale_x_continuous(breaks=seq(0, 130, 10))


# Semi-variogram

geoVane = as.geodata(blueVanes)

empGeor = variog(geoVane, max.dist=maxDist, breaks=seq(0, maxDist, by=6))
plot(empGeor)


```

Answer the question here.

The correlogram suggests that mean bee abundance does display spatial dependence up until approximately 20m between sampling points (traps). At that distance onward (from ~ 20m until 2/3 max distance ~ 89m), the effect size of any statistical correlation would be so low that there is likely no biological significance. The semi-variogram roughly supports this distance as the semivariance starts to "level out". Interesting, there is another rise in semivariance that corresponds to the second "dip" in the correlogram. I am dismissing this pattern as the correlogram shows us that the probablity of a biological significance in this second "dip"/ increase in semivariance is low.

# Challenge 3 (4 points)

Fit a simple linear model (assuming a normally distributed error term) that models mean bee abundance as a function of an intercept only (no other covariates in the model). Compare the model-estimated intercept to the bee abundance values averaged across all sites. What do you notice?  

Empirically test for evidence of spatial autocorrelation IN THE MODEL RESIDUALS by first building a correlogram, then building a semi-variogram. Be sure to provide an envelope for assessing whether the values in each figure are statistically different from what would be expected if the bee values were randomly distributed in space. Compare these figures to those built for the raw data. What similarities or differences to you notice?

```{r}
# Int only model

simMod <- lm(meanBee ~ 1, data = blueVanes)

# B0 vs average

summary(simMod) # B0 = 13.9322

mean(blueVanes$meanBee) # 13.9322

# Make df with residuals

BV2 <- data.frame(x = blueVanes$x, y = blueVanes$y, resid = simMod$residuals)

# Correlogram - MC permutation (if our values were randomly distributed)

correlogNcf2 = ncf::correlog(x = BV2$x, y = BV2$y, z = BV2$resid, increment=10, resamp=100)

tmp2 = cbind(correlogNcf2$n,
            correlogNcf2$mean.of.class,
            correlogNcf2$correlation,
            correlogNcf2$p) %>% 
  data.frame() %>% 
  set_names(c('n', 'dist', 'corr', 'p')) %>% 
  mutate(significant = ifelse(p < 0.05, 'yes', 'no'))

ggplot(tmp2, aes(x=dist, y=corr))+
  geom_line()+
  geom_point(aes(color=significant, size=n))+
  scale_color_manual(values=c('black', 'red'))+
  theme_bw()+
  theme(panel.grid=element_blank())+
  geom_hline(yintercept=0, linetype='dashed')+
  geom_vline(xintercept=maxDist, color='red')+
  scale_x_continuous(breaks=seq(0, 130, 10))

# Semi-variogram

geoVane2 = as.geodata(BV2)

empGeor2 = variog(geoVane2, max.dist=maxDist, breaks=seq(0, maxDist, by=5))
plot(empGeor2)

```

Answer the question here.

The model B0 is the same as the average of the values, which is to be expected. The correlograms and semi-variograms for the raw data and the residuals are also the same. Furthermore, if you look at the values in correlogNcf and correlogNcf2, they are also the same. I think that this makes sense because using the residuals from an intercept only model should essentially act like a data transformation. Essentially, all that we have done is made the new mean zero. We have not affected the actual variance of the data at all. Either that or I did this wrong, which is entirely possible.

# Challenge 4 (4 points)

Conventional wisdom indicates that the distribution of floral resources may substantially influence bee abundance. In addition, I am completely making up the following fact: the average bee has a perceptual range of about 15 m. Below I have provided code for importing a raster that shows the number of flowers counted in 1 m grid cells across the entire study area. Use the focal() function from the previous lab to generate a new raster in which each cell value represents the mean number of flowers in cells within a 15 m radius. Plot the distribution of the blue vane traps over the 15 m focal raster. No text required here, just show me the plot.

```{r}
# load rast

floral = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week6/floralResources.tif')

# prob matrix

probMatrix = focalMat(floral, 15, type='circle', fillNA=FALSE)

# I don't know if there are NAs

floralWith0 = floral
floralWith0[is.na(floralWith0)] = 0

# Summarize mean number of flowers in 15m radius

flor15m = focal(floralWith0, probMatrix, fun='mean')

# points on raster (I think this is what you want)

plot(flor15m)
points(blueVanes)

```

No text required

# Challenge 5 (4 points)

Extract the focal flower value from the raster you just created at each point using the extract() function. You now have an estimate for the average number of flowers per meter around each blue vane trap. Fit a second linear model, but this time include average floral abundance within 15 m as a covariate. Export the residuals from this fitted model, and again, examine a corelogram and variogram of the residuals. As always, be sure to plot random envelopes. Now what do you notice?

```{r}
# extract and make df

bvPoints <- data.frame(x = blueVanes$x, y = blueVanes$y)

rasValue=extract(flor15m, bvPoints)

bvFlo <- blueVanes
bvFlo$FloAbund15 <- rasValue$focal_mean

# Floral abundance model

floMod <- lm(meanBee ~ FloAbund15, data = bvFlo)
summary(floMod)

# Make df with residuals

bvFloResid <- data.frame(x = blueVanes$x, y = blueVanes$y, resid = floMod$residuals)

# Correlogram - MC permutation (if our values were randomly distributed)

correlogNcf3 = ncf::correlog(x = bvFloResid$x, y = bvFloResid$y, z = bvFloResid$resid, increment=10, resamp=100)

tmp3 = cbind(correlogNcf3$n,
            correlogNcf3$mean.of.class,
            correlogNcf3$correlation,
            correlogNcf3$p) %>% 
  data.frame() %>% 
  set_names(c('n', 'dist', 'corr', 'p')) %>% 
  mutate(significant = ifelse(p < 0.05, 'yes', 'no'))

ggplot(tmp3, aes(x=dist, y=corr))+
  geom_line()+
  geom_point(aes(color=significant, size=n))+
  scale_color_manual(values=c('black', 'red'))+
  theme_bw()+
  theme(panel.grid=element_blank())+
  geom_hline(yintercept=0, linetype='dashed')+
  geom_vline(xintercept=maxDist, color='red')+
  scale_x_continuous(breaks=seq(0, 130, 10))

# Semi-variogram

geoVane3 = as.geodata(bvFloResid)

empGeor3 = variog(geoVane3, max.dist=maxDist, breaks=seq(0, maxDist, by=5))
plot(empGeor3)

```

Answer the question here.

So, now the data does not appear to display spatial dependence via both the correlogram and semi-variogram. In the correlogram, we no longer see high correlation at any distance (and the axis scale has shrunk considerably). In the semi-variogram, the semi-variance is pretty standard across the entire plot, indicating that distance has no impact on residual variance. I think that this makes sense as we have measured the underlying covariate that accounted for the spatial dependence. Much like elevation from our in-class example, accounting for this previously "unknown" covariate has removed any spatial dependency in our data.

# Challenge 6 (3 points)

What is your assessment regarding the distribution of bee abundance in this study area? Does it exhibit spatial dependence, and if so, what is driving that pattern?

Answer the question here.

I kind of referenced this above, but no, I do not think that bee abundance actually exhibits spatial dependence in this area. Going from our discussion yesterday, if any spatial dependence in the raw data can be explained by a covariate, then it isn't truly spatial dependence in the response variable. It is simply an artifact of the patchy distribution of a prevalent covariate. Thus, our model including the covariate floral distribution eliminated the spatial dependence of the response variable, mean bee abundance. So, while the raw data implies that there is some degree of spatial dependance, I believe that the change in residuals after adding the covariate of floral distribution supports a lack of true spatial dependence in bee adundance and that the spatial dependence observed in the raw data is an artifact of the patchiness of the covariate. I do think that this could be a "slippery slope" in extreme circumstances such as data mining/dredging, but if there is a legitimate biological justification (as there is in bee abundance and floral distribution), then this concern is null. 
